General purpose gene annotation using deep learning

### 2020-04-20 paper/main.tex
text

\begin{methods}
\section{Methods}
We worked with 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 (?) \citealp{Howe19} as well as those of all non-embargoed plant genomes of the JGI Phytosome 13 database (?, citation). Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into training and test sets. The test sets contained 19 or 6 genomes, respectively. The test genomes were chosen to be of the best possible quality (?) based on collected metadata while also representing a broad phylogenetic spread. Some of the remaining genomes were used for training purposes and all were used for evaluating our models. We call the genomes that were ultimatily used during training the {\it training genomes} and the remaining non-test genomes the {\it validation genomes}.

We pre-processed and stored the genomic information by using Geenuff \citealp{Denton19}. Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. In the case of unambiguous input data, the genomic sequence was transformed into matrix that represents a one hot encoding with 4 rows. For the other case see Supplementary material .. (). The annotations are also transformed into a 4-row one hot encoding matrix, where a one in each of the rows represents one of the possible classes {\it Intergenic}, {\it UTR}, {\it CDS} or {\it Intron}. An non-coding intron can be inferred from the prediction order. A full description of all generated data arrays can be found in table \ref{tab:arrays}.

@@

\bibitem[Howe KL {\it et~al}., 2019]{Howe19}
Howe KL, Contreras-Moreira B, De Silva N, Maslen G, Akanni W, Allen J, Alvarez-Jarreta J, Barba M, Bolser DM, Cambell L, Carbajo M, Chakiachvili M, Christensen M, Cummins C, Cuzick A, Davis P, Fexova S, Gall A, George N, Gil L, Gupta P, Hammond-Kosack KE, Haskell E, Hunt SE, Jaiswal P, Janacek SH, Kersey PJ, Langridge N, Maheswari U, Maurel T, McDowall MD, Moore B, Muffato M, Naamati G, Naithani S, Olson A, Papatheodorou I, Patricio M, Paulini M, Pedro H, Perry E, Preece J, Rosello M, Russell M, Sitnik V, Staines DM, Stein J, Tello-Ruiz MK, Trevanion SJ, Urban M, Wei S, Ware D, Williams G, Yates AD, Flicek P. (2019) Ensembl Genomes 2020-enabling non-vertebrate genomic research. {\it Nucleic Acids Research 2019} https://doi.org/10.1093/nar/gkz890

>>>>

\begin{methods}
\section{Methods}
We worked with 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 (\citealp{Howe19}) as well as all non-embargoed plant genomes of the JGI Phytosome 13 database (\citealp{Phytozome}). Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into {\it training genomes} and {\it test genomes}. The test genome group is made up of 19 or 6 genomes, respectively, and was chosen to be of the best possible quality based on collected metadata while also representing a broad phylogenetic spread. We then experimented which genomes to use for during training from the pool of training genomes while evaluating the performance of models on the remaining ones. We call the genomes that were ultimatily not used during training the {\it validation genomes}. While we did split off a group of test genomes at the beginning of our work, we ended up only reporting on the generalization performance of the combined set of validation and test genomes as the validation genomes are by far the largest set. We also do not find a noticible difference in performance between those two groups.

We pre-processed and stored the genomic information by using Geenuff (\citealp{Denton19}). Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. In the case of unambiguous input data, the genomic sequence was transformed into matrix that represents a one hot encoding with 4 rows. For the other case see Supplementary material .. (). The annotations are also transformed into a 4-row one hot encoding matrix, where a one in each of the rows represents one of the possible classes {\it Intergenic}, {\it UTR}, {\it CDS} or {\it Intron}. An non-coding intron can be inferred from the prediction order. A full description of all generated data arrays can be found in table \ref{tab:arrays}.

@@

\bibitem[Howe {\it et~al}., 2019]{Howe19}
Howe KL, Contreras-Moreira B, De Silva N, Maslen G, Akanni W, Allen J, Alvarez-Jarreta J, Barba M, Bolser DM, Cambell L, Carbajo M, Chakiachvili M, Christensen M, Cummins C, Cuzick A, Davis P, Fexova S, Gall A, George N, Gil L, Gupta P, Hammond-Kosack KE, Haskell E, Hunt SE, Jaiswal P, Janacek SH, Kersey PJ, Langridge N, Maheswari U, Maurel T, McDowall MD, Moore B, Muffato M, Naamati G, Naithani S, Olson A, Papatheodorou I, Patricio M, Paulini M, Pedro H, Perry E, Preece J, Rosello M, Russell M, Sitnik V, Staines DM, Stein J, Tello-Ruiz MK, Trevanion SJ, Urban M, Wei S, Ware D, Williams G, Yates AD, Flicek P. (2019) Ensembl Genomes 2020-enabling non-vertebrate genomic research. {\it Nucleic Acids Research 2019} https://doi.org/10.1093/nar/gkz890

\bibitem[Goodstein {\it et~al}., 2012]{Phytozome}
	David M. Goodstein, Shengqiang Shu, Russell Howson, Rochak Neupane, Richard D. Hayes, Joni Fazo, Therese Mitros, William Dirks, Uffe Hellsten, Nicholas Putnam, Daniel S. Rokhsar (2012) Phytozome: a comparative platform for green plant genomics, {\it Nucleic Acids Research}, Volume 40, Issue D1, Pages D1178–D1186, https://doi.org/10.1093/nar/gkr944
